Interpretability and Fairness in Machine Learning: A Formal Methods Approach

### 2023-05-03 doctoral_consortium/main.tex
updated dc

% Comment out this line in the camera-ready submission
\linenumbers

@@

The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical, law, education, and transportation. In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users. With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2019stop}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox},  etc.\ in predictions. Towards responsible and trustworthy machine learning, we propose two research themes in our dissertation research: interpretability and fairness of machine learning classifiers. In particular, we design algorithms to \textit{learn interpretable rule-based classifiers}~\cite{ghosh22efficient,ghosh2019incremental,ghosh2020classification}, \textit{formally verify fairness}~\cite{ghosh2021justicia,ghosh2022algorithmic}, and \textit{explain the sources of unfairness}~\cite{ghosh2022how}. Prior approaches to these problems are often limited by scalability, accuracy, or both. To overcome these limitations, we closely integrate automated reasoning, formal methods, and statistics with fairness and interpretability to develop scalable and accurate solutions.

@@

Interpretable machine learning often employs rule-based classifiers, which use a set of rules to represent the decision boundary. The interpretability of such classifiers depends on the size of the rules: smaller rules with higher accuracy are preferred in practice. However, this presents a challenge when dealing with large datasets, as interpretable classification learning becomes a combinatorial optimization problem that suffers from poor scalability. To address this issue, we propose an incremental learning framework for interpretable rule-based classification on large datasets. Our framework combines maximum satisfiability (MaxSAT) and mixed integer linear programming (MILP) with mini-batch learning.

@@

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to enable the learning of a more relaxed representation of classification rules with higher expressiveness, as described in~\cite{ghosh2020classification}. Specifically, we consider relaxed definitions of the standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and in the selection of clauses in a rule. Based on these relaxed definitions, we introduce relaxed logical classification rules, which are motivated by the use of checklists in the medical domain and Boolean cardinality constraints. These rules generalize widely used rule representations, such as CNF, DNF, and decision sets. However, the combinatorial structure of these rules results in exponential succinctness, and na\"ive learning techniques are computationally expensive. To overcome this issue, we propose an incremental mini-batch learning procedure, called $\mathsf{CRR}$, which leverages advances in MILP solvers to efficiently learn such rules. Our experimental analysis shows that $\mathsf{CRR}$ can generate more accurate and sparser classification rules compared to alternative rule-based classifiers.

@@

\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness.
Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias. To understand the sources of bias, it's necessary to determine \textit{which factors contribute how much to the bias of a classifier on a dataset}. We use a feature-attribution approach to explain the sources of bias, which relates the influences of input features to the resulting bias of the classifier. We formalize the Fairness Influence Function (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}. We build an algorithm called $\mathsf{FairXplainer}$, which estimates FIFs by decomposing the variance of the classifier's prediction among all subsets of features, using global sensitivity analysis. In experiments, $\mathsf{FairXplainer}$ captures the influences of both individual and intersectional features across various datasets and classifiers, approximates bias more accurately using FIFs than existing local explanation methods, and demonstrates a higher correlation of FIFs with fairness interventions. In future, we aim to develop improved fairness enhancing and attack algorithms based on fairness verifier $ \mathsf{Justicia} $ and fairness explainer $ \mathsf{FairXplainer} $.

>>>>

% Comment out this line in the camera-ready submission
%\linenumbers

@@

The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical, law, education, and transportation. In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users. With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2022interpretable}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox},  etc.\ in predictions. Towards responsible and trustworthy machine learning, we propose two research themes in our dissertation research: interpretability and fairness of machine learning classifiers. In particular, we design algorithms to \textit{learn interpretable rule-based classifiers}~\cite{ghosh22efficient,ghosh2019incremental,ghosh2020classification}, \textit{formally verify fairness}~\cite{ghosh2021justicia,ghosh2022algorithmic}, and \textit{explain the sources of unfairness}~\cite{ghosh2022how}. Prior approaches to these problems are often limited by scalability, accuracy, or both. To overcome these limitations, we closely integrate automated reasoning, formal methods, and statistics with fairness and interpretability to develop scalable and accurate solutions.

@@

Interpretable machine learning often employs \textit{rule-based classifiers}, which use a set of rules to represent the decision boundary. One particular notion of interpretability of classification rules is their rule-size: smaller rules with higher accuracy are preferred in medical domains. However, this presents a challenge when dealing with large datasets, as interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability. To address this issue, we propose an incremental learning framework for rule-based classification on large datasets. Our framework combines maximum satisfiability (MaxSAT) and mixed integer linear programming (MILP) with mini-batch learning.

@@

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to enable the learning of a more relaxed representation of classification rules with higher expressiveness~\cite{ghosh2020classification}. Motivated by satisfying $ M $ out of $ N $ literals, we consider relaxed definitions of OR/AND operators in logic, which allows exceptions in the construction of a clause and in the selection of clauses in a rule. Based on these relaxed definitions, we introduce relaxed logical classification rules, which are inspired by the use of checklists in the medical domain and Boolean cardinality constraints. These rules generalize widely used rule representations, such as CNF, DNF, and decision sets. However, the combinatorial structure of these rules results in exponential succinctness, and na\"ive learning techniques are computationally expensive. To overcome this issue, we propose an incremental mini-batch learning procedure, called $\mathsf{CRR}$, which leverages advances in MILP solvers to efficiently learn such rules. Our experimental analysis shows that $\mathsf{CRR}$ can generate more accurate and sparser classification rules compared to alternative rule-based classifiers.

@@

\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness \textit{essentially beyond sensitive features}. Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias. To understand the sources of bias, it's necessary to determine \textit{which factors contribute how much to the bias of a classifier on a dataset}. We use a feature-attribution approach to explain the sources of bias, which relates the influences of input features to the resulting bias of the classifier. We formalize the Fairness Influence Function (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}. We build an algorithm called $\mathsf{FairXplainer}$, which estimates FIFs by decomposing the variance of the classifier's prediction among all subsets of features, using global sensitivity analysis. In experiments, $\mathsf{FairXplainer}$ captures the influences of both individual and intersectional features across various datasets and classifiers, approximates bias more accurately using FIFs than existing local explanation methods, and demonstrates a higher correlation of FIFs with fairness interventions. In future, we aim to develop improved fairness enhancing and fairness attack algorithms based on fairness verifier $ \mathsf{Justicia} $ and fairness explainer $ \mathsf{FairXplainer} $.
